\section{Geometry, Topology and Sets}

This section covers the fundamental mathematical toolset that future material depends on.
To begin with, affine and convex sets, convex functions and separation theorems are covered.
Following this, some topological concepts are discussed such as the closure of convex sets and functions, as well as the relative interior of a set.

\subsection{Affine and Convex sets}

The central object in section 1 of \cite{rockafellar-1970a} is the \textit{Affine set}.
\begin{Def}[Affine set]\cite[p.~3]{rockafellar-1970a}
    A subset $M$ of $\R^n$ is called an affine set if $(1 - \lambda)x + \lambda y \in M$ for every $x \in M$, $y \in M$ and $\lambda \in \R$
    \label{def:affine-set}
\end{Def}

An exercise left for the reader is to prove that the \textit{affine hull} of an arbitrary subset $S \subset \R^n$ consists of all vectors of the form $\lambda_1 x_1 + \dots + \lambda_m x_m$, such that $x_i \in S$ and $\lambda_1 + \dots + \lambda_m = 1$.
To prove this we begin with a definition.
\begin{Def}[Affine hull]
    The affine hull of an arbitrary subset $S \subset \R^n$ is the unique, smallest affine set that contains $S$. 
    In other words, the affine hull is the intersection of all affine sets containing $S$.
    We denote the affine hull of $S$ as $\operatorname{aff} S$.
\end{Def}

We begin by defining the set $C$ as all linear combinations of points in the arbitrary subset $S \subset \R^n$ with coefficients summing to one
\[
    C = \left\{ \sum_{i = 1}^m \lambda_i x_i \> | \> m \geq 1, x_i \in S, \sum_{i = 1}^m \lambda_i = 1  \right\},
\]
as well as stating, and proving, the following lemma.
\begin{Lma}
  Let $M \subseteq \mathbb{R}^n$ be affine. If $x_1,\dots,x_m \in M$ and
  $\lambda_1+\dots+\lambda_m = 1$, then
  \[
      \sum_{i=1}^m \lambda_i x_i \in M.
  \]
  \end{Lma}

  \textit{Proof}. We argue by induction on $m$.

  For $m=1$, the claim is immediate, since $\lambda_1 = 1$ and therefore
  \[
      \lambda_1 x_1 = x_1 \in M.
  \]

  For $m=2$, the claim follows directly from the definition of an affine set:
  if $\lambda_1+\lambda_2=1$, then $\lambda_1 = 1-\lambda_2$, so
  \[
      \lambda_1 x_1 + \lambda_2 x_2
      =
      (1-\lambda_2)x_1 + \lambda_2 x_2
      \in M.
  \]

  Now assume the result holds for affine combinations of $m-1$ points, and
  let
  \[
      \sum_{i=1}^m \lambda_i x_i
  \]
  be an affine combination of $m$ points in $M$.

  Since $m \geq 2$, at least one coefficient is different from $1$.
  After relabeling if necessary, assume $\lambda_m \neq 1$. Then
  \[
      \sum_{i=1}^m \lambda_i x_i
      =
      (1-\lambda_m)
      \left(
          \sum_{i=1}^{m-1}
          \frac{\lambda_i}{1-\lambda_m}x_i
      \right)
      +
      \lambda_m x_m.
  \]
  The coefficients inside the parentheses sum to
  \[
      \sum_{i=1}^{m-1}
      \frac{\lambda_i}{1-\lambda_m}
      =
      \frac{1-\lambda_m}{1-\lambda_m}
      =
      1.
  \]
  Hence, by the induction hypothesis,
  \[
      \sum_{i=1}^{m-1}
      \frac{\lambda_i}{1-\lambda_m}x_i
      \in M.
  \]
  Since also $x_m \in M$ and $M$ is affine, their affine combination belongs
  to $M$. Therefore
  \[
      \sum_{i=1}^m \lambda_i x_i \in M.
  \]
  This completes the induction. \hfill\sq

With these bits in mind we are ready to prove what is stated in the book which we formulate as the following theorem.
\begin{Thm}
    $\operatorname{aff} S = C$
\end{Thm}
\textit{Proof}. The proof is divided into two parts, showing $C \subseteq \operatorname{aff} S$ and showing $\operatorname{aff} S \subseteq C$, which together imply the statement of the theorem.
\begin{enumerate}
    \item[Case $C \subseteq \operatorname{aff} S$: ] The lemma above implies any linear combination of points with coefficients summing to one is a valid affine combination. Thus, since the affine hull contains $S$, the elements of $C$ are affine combinations of elements in the affine hull and thus $C\subseteq \operatorname{aff}S$. 
    \item[Case $\operatorname{aff} S \subseteq C$] It remains to show that $\operatorname{aff} S \subseteq C$.
  Since $C$ contains $S$, it suffices to show that $C$ is affine. Indeed, then $C$ is one of the affine sets containing $S$, and since $\operatorname{aff} S$ is the smallest such set, we get $\operatorname{aff} S \subseteq C$.
    Take two arbitrary points $x, y \in C$. These can be written as
    \begin{align*}
        x &= \sum_{i = 1}^{m_x} \lambda_i x_i, \quad\quad \sum_{i = 1}^{m_x} \lambda_i = 1\\
        y &= \sum_{i = 1}^{m_y} \gamma_i y_i, \quad\quad \sum_{i = 1}^{m_y} \gamma_i = 1
    \end{align*}
    We form an affine combination of $x$ and $y$ to see if $C$ is closed under affine combinations,
    \[
        z = ax + (1 - a)y = \sum_{i = 1}^{m_x} a\lambda_i x_i + \sum_{i = 1}^{m_y} (1 - a) \gamma_i y_i 
    \]
    This is a linear combination of points in $S$ as $x_i, y_i \in S$ with coefficients $a\lambda_i$ and $(1-a)\gamma_i$.
    To show this is in $C$ we only have to show all combinations sum to one.
    \begin{align*}
        1 &\overset{!}{=} \sum_{i = 1}^{m_x}a\lambda_i + \sum_{i = 1}^{m_y}(1-a)\gamma_i \\
          &= a\sum_{i = 1}^{m_x}\lambda_i + (1 - a)\sum_{i = 1}^{m_y}\gamma_i\\
          &= a + (1-a) = 1
    \end{align*}
    Thus $C$ is an affine set containing $S$ which implies $\operatorname{aff} S \subseteq C$ and the proof is complete. \hfill\sq
\end{enumerate}

\subsection{Cones}

A subset $K \subseteq \R^n$ is called a cone if it is closed under positive scalar multiplication, i.e. $\lambda x \in K$ if $\lambda > 0$ and $x \in K$.
Naturally there is such a thing as a \textit{convex cone}.
That is a cone which is also a convex set.
When reading this at first, I thought cones were almost always convex and couldn't really imagine a cone which wasn't convex.
The mental, geometrical image I had of a cone was of the ``world sheet'' swept out of an expanding circle emanating from the origin which is convex, no doubt.
But it turns out, cones can look very different indeed, and the following example of a non-convex cone illuminates
\[
    K_{nc} = \{(x, 0)\> |\> x \geq 0\} \cup \{(0, y)\> |\> y \geq 0\} 
\] 

\input{figures/non-convex-cone.tex}

It is definitely closed under positive scalar multiplication, but not under convex combinations. 
For example $(1/2)((1, 0) + (0, 1)) = (1/2, 1/2) \notin K_{nc}$.

\subsection{Solved exercises}

In this section are solutions to some exercises testing foundational understanding of convexity.
The first set of exercises from ''Convex Optimization`` are from the first homework in the Stanford course EE364a.

\subsubsection{CO: 2.5}
\textbf{What is the distance between two parallel hyperplanes $P_1 = \{x \> | \> a^T x = b_1\}$, $P_2 = \{x \> | \> a^T x = b_2\}$?}\\
The distance can be found by taking the difference between two points in the planes and looking at the component perpendicular to the planes.
The vector $a$ is normal to the plane, so by computing the unit vector in that direction we get a vector with which we can obtain the perpendicular component.
So, if we take $x_1 \in P_1$, $x_2 \in P_2$, we get the distance as follows:
\[
    d = \frac{|a^T(x_2 - x_1)|}{\|a\|_2} = \frac{|b_2 - b_1|}{\| a \|_2}
\]

\input{figures/parallel-plane-distance.tex}

\subsubsection{CO: 2.7}
\textbf{Voronoi description of halfspace. Let $a$ and $b$ be distinct points in $\mathbb{R}^n$. Show that the set
of all points that are closer (in Euclidean norm) to $a$ than $b$, i.e., $\{x | \|x-a\|_2 \leq \|x-b\|_2\}$,
is a halfspace. Describe it explicitly as an inequality of the form $c^T x \leq d$. Draw a picture.}

The situation to be imagined is one with two points separated by a hyperplane perpendicular to the line connecting the two points, placed exactly at the midpoint of the connecting line segment, see fig \ref{fig:voronoi-halfspace}.
The line passing through the points is generated by the vector $a-b$ which is naturally also the normal of the separating hyperplane.
The separating hyperplane defines two halfspaces, where all points closer to $a$ than $b$ are in the halfspace containing $a$.
Since $a-b$ points from $b$ to $a$, it means that points with $(a - b)^T x$ larger than at the midpoint are in the halfspace containing $a$.
Thus, we only need to compute the innerproduct of $(a - b)$ and the midpoint between the points, $(a + b) / 2$.

\input{figures/voronoi-halfspace.tex}

\[
    d = \frac{(a - b)^T(a + b)}{2} = \frac{a^Ta - b^Tb}{2}
\]

Thus, all points $x$ for which $(a-b)^T x$ is larger than $d$ are closer to $a$ than $b$.
The description of the halfspace as sought in the question is
\[
    \{x | \|x-a\|_2 \leq \|x-b\|_2\}
    = \{x | (a-b)^T x \geq d\}
    = \{x | (b-a)^T x \leq -d\},
\]
with $c = b - a$, and $-d = (b^Tb - a^Ta)/2$.
